\documentclass[12pt]{article}

% House style preamble
\input{.house-style/preamble.tex}

% Update PDF metadata
\hypersetup{
  pdftitle={A metaphysician's systemic-functional turn: Thomasson's Rethinking Metaphysics and the architecture of linguistic description},
  pdfkeywords={systemic functional linguistics, conceptual engineering, linguistic description, grammatical architecture, lexical category, syntactic function, discourse role, grammatical metaphor, metaphilosophy}
}

% Working title -- provisional, confirm with Brett.
\title{A metaphysician's systemic-functional turn\\
{\large Amie Thomasson's \emph{Rethinking Metaphysics} and the architecture of linguistic description}}
\author{Brett Reynolds \orcidlink{0000-0003-0073-7195}%
\thanks{Contact: \href{mailto:brett.reynolds@humber.ca}{brett.reynolds@humber.ca}}\\
Humber Polytechnic \& University of Toronto}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
\posscite{thomasson2025rethink} \emph{Rethinking Metaphysics} brings a grammatical debate into metaphysics. Thomasson uses systemic functional linguistics to explain why noun-term recastings such as \mention{the property of redness} are useful without behaving like observational object talk. This paper accepts the discourse-role insight but detaches it from Halliday's grammatical architecture. SFL explains why the recast formulations are worth having; easy ontology licenses the inferences. For linguists, the lesson is broader than syntax: some inferences and expectations follow from linguistic description, but not every expectation imported from familiar models of reference, objecthood, or category.
\end{abstract}

\noindent\textbf{Keywords:} systemic functional linguistics; conceptual engineering; linguistic description; grammatical architecture; lexical category; syntactic function; discourse role; grammatical metaphor; metaphilosophy

\section{Introduction}
What's surprising about \posscite{thomasson2025rethink} \emph{Rethinking Metaphysics} is that the constructive half of the book is organized around a Hallidayan way of asking what language is for. Thomasson turns to systemic functional linguistics to explain why easy ontological inferences can be ordinary linguistic moves, why metaphysical disputes so often misfire, and how conceptual engineering should begin.

That should matter to linguists. A debate internal to grammar~-- whether discourse roles organize grammatical description or are better kept distinct from syntactic category and syntactic function~-- has become part of a live metaphilosophical project. This paper's claim is correspondingly double. Thomasson is right that linguistic roles matter for metaphysics. Her best results, though, don't require the systemic-functional placement of those roles inside grammatical architecture.

The aim is to extract the part of SFL Thomasson needs, not to dismiss the borrowing. That part is detachable from Halliday's stronger architectural commitments. The autonomy claim is minimal: discourse roles are real and explanatory, but they don't determine syntactic category or syntactic function. I focus on noun-term recastings because they make that issue clearest. Modal vocabulary remains part of the Thomassonian background, not a second case study.

\section{Thomasson's systemic-functional turn}
Part I of \emph{Rethinking Metaphysics} has a diagnostic role in the book's metametaphysical argument. It asks whether familiar projects can supply a uniform standard for legitimate metaphysical discourse: joint-carving, grounding, truthmaking, causal access, or explanatory contribution. Thomasson's answer is no. These projects look for the wrong kind of standard. The constructive half changes the question. Instead of asking what all metaphysical discourse has to do, it asks what different linguistic resources let speakers do, how they entered the language, and whether they should be retained, revised, or rejected.

Systemic functional linguistics enters here. Thomasson says that SFL gives her a way to identify discourse roles and explain why philosophically suspicious expressions can still be in good order \citep[133--160]{thomasson2025rethink}. Her central examples are familiar from easy ontology: if the barn is red, then there's a property, redness; if there are two cups on the table, then the number of cups on the table is two \citep[169--179]{thomasson2025rethink}. A critic hears these transitions as cheap routes to abstract objects. Thomasson hears them as legitimate linguistic moves, useful because of what the derived formulations let speakers do.

Halliday's background claim isn't just that language is useful. In Halliday and Matthiessen's statement of the theory, language simultaneously construes experience, enacts social relations, and organizes discourse. The three modes are the ideational, interpersonal, and textual metafunctions. They're called \term{metafunctions} because, in the theory, function is internal to linguistic architecture: \enquote{the entire architecture of language is arranged along functional lines} \citep[30--31]{hallidayMatthiessen2014introduction}. That's the architectural claim at issue in this paper.

Thomasson uses the architecture most directly through what she, following Halliday, calls \term{grammatical metaphor}. In Hallidayan terms, a meaning has a more congruent wording and can also be construed through another grammatical form. Thomasson stresses that the resulting claims aren't fictional or merely figurative: grammatical metaphor is a way mature language extends what speakers can literally say \citep[154--156]{thomasson2025rethink}.

In the easy-ontology cases, the relevant grammatical metaphors are NP-forming recastings that introduce noun terms. \mention{The barn is red} supports \mention{The barn possesses} [\mention{the property of redness}]; \mention{There are two cups on the table} supports [\mention{The number of cups on the table}] \mention{is two}; \mention{May was born on a Monday} supports [\mention{The event of May's birth}] \mention{occurred on a Monday} \citep[169--179]{thomasson2025rethink}.

I call these local cases \term{noun-term recastings}: NP-forming recastings that introduce terms for reference, predication, comparison, and theory-building. For Thomasson, abstract-object vocabulary misleads when philosophers carry over expectations from ordinary object NPs such as \mention{a barn} and \mention{a cup}. The bracketed NPs are grammatically ordinary, but their discourse role is to package qualities, cardinalities, and events. They don't have to introduce objects on the model of observable things.

SFL helps Thomasson move from negative deflationary arguments to a positive method for metaphysics as \term{conceptual engineering}. Here that means assessing and sometimes revising linguistic resources: what they let speakers do, whether they mislead, and which uses are worth retaining. Her method asks us to identify an expression and the congruent or base formulation against which it's analyzed, then ask what needs that base formulation served and what the incongruent formulation adds \citep[203--207]{thomasson2025rethink}. That substantive use of linguistic theory in philosophy calls for linguistic assessment.

Thomasson's own shift from grammatical form to philosophical payoff makes the risk visible. She describes the SFL part of the project as a \enquote{formal} assessment of function, one concerned with functions \enquote{based in the formal-grammatical forms employed} \citep[206]{thomasson2025rethink}. She then asks what's added by forms such as \mention{there is a possibility of rain} rather than \mention{it might rain}, or by noun forms for \mention{knowledge} and \mention{truth} rather than clausal formulations \citep[206--207]{thomasson2025rethink}. The grammatical contrast is real. The question is whether the formal-grammatical description fixes the discourse role.

\section{Function, role, and placement}
The word \term{function} creates a risk of slippage. I use three labels for three levels. A \term{syntactic function} is a structural relation inside a construction. A Hallidayan \term{metafunction} is a theory-internal organizing dimension of grammar. A \term{discourse role} is what an expression lets speakers do in a practice. These labels mark different explanatory questions, not one quantity.

For the rest of the paper, I avoid bare \term{function} except when naming the ambiguity or reporting Hallidayan terminology. The paper's criticism isn't that discourse roles are unreal. It's that the third item shouldn't be collapsed into either of the first two.

In the first sense, familiar from \emph{The Cambridge Grammar of the English Language} (\emph{CGEL}) and its descendants, a syntactic function is a relational position in constructional structure: \term{subject}, \term{predicator}, \term{determiner}, \term{head}, \term{modifier}, \term{complement} \citep{huddleston2002}. Syntactic function differs from lexical and phrasal category.

In \mention{A stone cracked}, \mention{stone} is a \term{noun} heading the \term{NP} \mention{a stone}, which functions as \term{subject} of the clause. In \mention{stone wall}, \mention{stone} heads a \term{Nom} functioning as \term{modifier} within a larger \term{Nom}. The same \term{modifier} function can also be filled by an \term{AdjP}: in \mention{red wall}, \mention{red} heads an \term{AdjP} functioning as \term{modifier}; in \mention{The wall is red}, the \term{AdjP} functions as \term{predicative complement}.

In the second sense, \term{functional} is Hallidayan. Language is described in terms of what it enables speakers to do, but that claim is built into grammatical architecture. Halliday's metafunctions~-- ideational, interpersonal, and textual~-- organize the grammar itself, rather than entering as later pragmatic uses of an independently described syntax. Thomasson adopts this picture when she turns to Hallidayan linguistics for a systematic account of what she calls linguistic functions \citep[139--159]{thomasson2025rethink}.

On that analysis, \mention{The barn is red} is simultaneously an ideational relational clause, with \mention{the barn} as carrier and \mention{red} as attribute; an interpersonal mood structure, with \mention{the barn} as Subject and \mention{is} as Finite; and a textual structure, with \mention{the barn} as Theme and \mention{is red} as Rheme.

In the third sense, a discourse role is a pragmatic or teleological role in a practice. Thomasson has this sense in view when she asks what a stretch of discourse does for speakers, why it persists, and whether it should be retained, revised, or abandoned. Syntactically interrogative clauses show the same separation: \mention{Did Kim leave?} asks a question; \mention{Can you pass the salt?} makes a request; \mention{Would you like tea?} makes an offer; and \mention{Why not take the train?} makes a suggestion.

Thomasson's metaphysical examples make the third sense more explicit. In \mention{The barn has the property of redness}, the relevant NP is \mention{the property of redness}. Its discourse role isn't naming a perceptual object. The NP packages a quality so speakers can qualify it, compare it, refer back to it, and build theory with it.

For conceptual engineering, the third sense matters most. A reverse-engineering project asks what an expression is good for before deciding whether it's philosophically suspect \citep[203--207]{thomasson2025rethink}. That role can be descriptive or normative: what the expression has done, what it now does, or what we'd endorse keeping it for.

Discourse roles matter. The question is where a grammatical theory should locate them. A distribution-first grammar can agree that noun-term recastings let speakers quantify, qualify, package information, and build theories. It can also agree that modal language serves social and normative ends.

These roles may belong in a broader account of the inferences and uses supported by a construction. The open question is whether those roles are category-making conditions for syntactic category or syntactic function. The roles are real; the placement is optional.

Thomasson's borrowing becomes linguistically interesting at this point. For philosophers, systemic functional linguistics offers an attractive corrective to representational monism: many expressions earn their keep without naming or tracking an object. For an autonomist syntactician, the corrective comes bundled with an architectural choice. Halliday is right to insist that expressions serve multiple discourse roles. On this view, the problematic step is treating those roles as part of the architecture of grammatical description rather than as roles made available by grammatically describable forms.

That last claim is about explanatory placement, not about whether SFL is useful. A grammar can identify the relevant expressions and describe the constructions that make discourse roles available; a broader discourse account explains why those roles matter.

The rejected step isn't including those roles in a broader account of supported inferences and uses. It's treating them as what fixes lexical category or syntactic function: whether a form belongs to the category \term{noun} or \term{determinative}, or whether an expression has \term{subject} or \term{determiner} function. The \emph{CGEL}-style objection doesn't say Halliday is wrong to notice that clauses, phrases, and NPs can do similar discourse work. It says shared discourse work shouldn't be recoded as syntactic category.

\section{Grammatical metaphor and the category/discourse-role confound}
Grammatical metaphor is the pressure point. Thomasson is careful to block a bad philosophical reading of the term. In her Hallidayan use, \term{grammatical metaphor} names a relation between congruent and incongruent grammatical realizations.

The word \term{congruent} depends on the SFL architecture already introduced. Halliday treats lexicogrammar as the stratum that realizes metafunctionally organized meaning. In that architecture, a congruent realization follows the theory's default mapping from metafunctionally organized meaning to grammatical resource. On the ideational side, processes are realized as verbal groups, qualities as adjectival resources, and participants as nominal groups. On the interpersonal side, questions are realized in interrogative mood and commands in imperative mood. The theory's architecture supplies that expectation, with developmental and historical arguments offered in support \citep[659--665, 707--713]{hallidayMatthiessen2014introduction}.

That source matters. If the congruent/incongruent contrast is treated as neutral description, the architectural decision disappears from view. The SFL vocabulary can then make a theory-internal mapping look like a fact supplied by grammar itself. A distribution-first grammar will state the relation more cautiously: some recastings are useful because they make new discourse work available, but that use doesn't settle syntactic category or syntactic function.

A distribution-first grammar proceeds in the other direction. It identifies [\mention{the property of redness}], [\mention{the number of cups on the table}], and [\mention{the event of May's birth}] as NPs because of their distribution: they occur in NP positions, include determiners and heads such as \mention{property}, \mention{number}, or \mention{event}, take ordinary complements and modifiers, and enter into NP-level dependencies. Their discourse roles are then described separately. They package qualities, cardinalities, and events for reference, comparison, anaphora, and theoretical elaboration. The first fact is syntactic; the second is discourse-functional. Neither reduces to the other.

Among the easy-ontology examples, the cases that matter here are noun-term recastings: an adjectival quality-ascription can be recast with [\mention{the property of redness}]; the cups case shifts from a cardinality claim to a number-ascription, from \mention{There are two cups on the table} to [\mention{The number of cups on the table}] \mention{is two}; an event description can support [\mention{the event of May's birth}] \citep[154, 169--179]{thomasson2025rethink}. The resulting claims aren't make-believe. Mature language lets speakers recast material across grammatical forms.

These noun-term shifts really do add capacities \citep[155--156]{thomasson2025rethink}. They let speakers qualify and quantify over the recast material, increase lexical density, package earlier claims, and construct layered theories. This lets Thomasson say that systemic functional linguistics gives her a new answer to a question left open by easy ontology: why would a language contain apparently redundant transitions from \mention{the barn is red} to \mention{the barn has the property of redness}? Such transitions are useful. They let speakers do work that the base formulations don't do as well.

But usefulness isn't validity. Thomasson had already defended the easy inferences by appeal to introduction rules, redundancy, and ordinary linguistic competence before the systemic-functional apparatus enters the 2025 book \citep[34--39]{thomasson2025rethink}. She also says that, when she wrote \emph{Norms and Necessity}, she hadn't yet discovered systemic functional linguistics \citep[xiv]{thomasson2025rethink}; the modal-normativist program was already in place \citep{thomasson2020norms}. The SFL material helps explain why the NP formulations are worth having. By itself, it supplies no rule that licenses the inference.

This distinction matters because the SFL vocabulary can make a category contrast look like a discourse-role explanation. In the property case, the contrast is between adjectival and NP formulation. In \emph{CGEL} terms, that's a contrast in category and construction. The pragmatic payoff is a further fact about what the NP construction lets speakers do. If we collapse the two, we risk treating an expression's usefulness as if it explained its syntactic status, or treating a syntactic category assignment as if it by itself justified an ontological inference.

Thomasson's own later discussion pushes in the same direction. In Chapter 9, she says that the formal-functional analysis is incomplete on its own. Knowing that modal expressions or noun-term recastings serve broad formal roles doesn't yet distinguish the discourse roles of different modal flavors or different grammatical metaphors. For that, she adds a needs-based, material analysis: identify the base formulation, then ask what needs or desires it serves and what the corresponding grammatical metaphor adds \citep[204--207]{thomasson2025rethink}. That later move is evidence that grammatical architecture underdetermines the philosophically relevant role distinctions.

Keep the discourse-role turn. Thomasson's strongest conclusions survive a less Hallidayan grammar. She needs the claim that expressions have many roles, that noun-term recastings expand what speakers can do, and that treating noun-term formulations on the model of observational object tracking produces pseudo-problems. The stronger claim that those roles belong inside grammatical architecture is dispensable. Halliday gives her a powerful diagnostic vocabulary, but the inferential work is done by easy ontology plus framework-neutral facts about what the resulting NPs let speakers do.

\section{What linguists should make of it}
First, linguistic theories travel. When they travel into philosophy, education, law, or cognitive science, their internal disputes often disappear from view. Thomasson's reader may come away thinking that Hallidayan grammar is simply what linguistics says once it takes discourse roles seriously. That would be a mistake. It's one specific theoretical tradition within linguistics, not the neutral name for role-sensitive grammatical explanation.

Second, autonomist grammarians shouldn't answer Thomasson by denying the discourse-role facts. Noun-term recastings let speakers package qualities, cardinalities, and events as NPs, so that speakers can quantify over, modify, specify, and develop the resulting material theoretically. Modal vocabulary does do normative and social work. The better answer is to separate those facts from the claim that the grammar itself has to be organized by the roles. The autonomy of syntax isn't the claim that language has no uses. It's the claim that syntactic description shouldn't be made hostage to a prior inventory of discourse roles.

Third, Thomasson's examples expose an unlicensed carry-over: an NP needn't support the same ontological inferences merely because it's an NP. A syntactic-category assignment carries expectations about distribution and dependencies. The metaphysical question concerns a different target: which uses of NPs, especially NPs introduced by noun-term recastings, support which existential, explanatory, or practical inferences, under which conditions. That broader question belongs to a larger philosophical project. The narrower claim here is that syntactic category, syntactic function, and discourse role answer different questions.

For linguists, then, \emph{Rethinking Metaphysics} is both an opportunity and a warning. The opportunity is to show philosophers that detailed grammatical distinctions matter for live metaphysical questions. The warning is that, when lexical category, syntactic function, and discourse role stay implicit, SFL's architecture can look like part of the linguistic result rather than one theory's way of framing it. Thomasson needed a way to say that abstract-object talk can be useful without being observational object tracking. SFL gave her an attractive vocabulary for that task. A clearer grammar can keep the insight and remove the unnecessary commitment.

\section{Conclusion}
Thomasson's systemic-functional turn is a real event in the afterlife of linguistic theory, but not because SFL has rebuilt metaphysics.

A major metaphysical project reaches for a linguistic vocabulary at just the point where the validity of easy inferences is no longer the only question. Thomasson also needs to say what the resulting formulations are for, why the recast NPs are worth having, and why treating them on the model of observational object tracking misdescribes them. SFL supplies an attractive vocabulary for that task. The attraction shouldn't be mistaken for dependence. The grammatical-metaphor apparatus explains the usefulness of noun-term recasting; the philosophical licensing comes from elsewhere.

For linguists, that makes \emph{Rethinking Metaphysics} more than a case of exported theory. It shows why metaphysical uses of linguistic description are worth taking seriously. The worked case starts from syntactic category: the ordinary NP status of [\mention{the property of redness}], [\mention{the number of cups on the table}], and [\mention{the event of May's birth}]. But the lesson is broader. Thomasson's book presses linguists to ask what our descriptions are descriptions of, what inferences they support, and when those inferences have been carried over from familiar models of reference, objecthood, or category rather than earned by the analysis.

Semanticists can ask what follows from reference, predication, or denotation; discourse theorists can ask what follows from packaging, anaphora, uptake, or theoretical elaboration. The same question recurs wherever a linguistic analysis is used to support expectations beyond the description itself.

Neither dismissal nor conversion follows. Thomasson is right to reject a one-role model of language, and right that many metaphysical puzzles arise when a tracking model is extended across expressions with different roles. Her case also teaches linguistic theorists something: metaphysical disputes can show which expectations follow from linguistic description and which have been imported from a familiar concrete-object noun model.

A linguistic account of those results needs a grammar that keeps syntactic distribution, syntactic category, syntactic function, and discourse role apart while asking what follows from each. Here, metaphysics becomes useful for linguistics: not as a source of grammatical categories, but as a test of what follows from them.

\section*{Acknowledgements}

OpenAI Codex and Anthropic Claude Code served as drafting and editing aids during the preparation of this paper. I am responsible for all theoretical claims, arguments, errors, and interpretive choices.

\clearpage
\printbibliography

\end{document}
